---
title: "ggplot2 컬러링 가이드"
subtitle: "서울 R 미트업"
author: "방태모"
institute: "가천대 길병원 인공지능빅데이터융합센터"
date: "2022-08-12"
output:
  xaringan::moon_reader:
    css: 
      - default
      - css/statisticsplaybook.css
      - css/statisticsplaybook-fonts.css
    lib_dir: libs
    seal: false
    nature:
      highlightStyle: googlecode
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
          scroll: false # disable slide transitions by scrolling
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
      after_body: [css/insert-logo.html]
---

```{r setup, include = FALSE}
library(knitr)
library(rmarkdown)
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
# set default options
# opts_chunk$set(dpi = 300)
```

class: title-slide, left, bottom

# `r rmarkdown::metadata$title`
----
## **`r rmarkdown::metadata$subtitle`**
### `r rmarkdown::metadata$author`
### `r rmarkdown::metadata$date`

---

## .brand-blue[준비하기]


```{r, message = FALSE, warning=FALSE}
library(tidyverse)
library(showtext)
font_add_google("Fira Sans", "firasans")
showtext_auto()

theme_customs <- theme(
  text = element_text(family = 'firasans', size = 16),
  plot.title.position = 'plot',
  plot.title = element_text(
    face = 'bold', 
    colour = thematic::okabe_ito(8)[6],
    margin = margin(t = 2, r = 0, b = 7, l = 0, unit = "mm")
  )
)

theme_set(theme_minimal() + theme_customs)
```

---

class: inverse, middle, center

# 음영을 활용하자

---

## .brand-blue[막대그래프 예제]

- 1998년과 2008년 인기 차량 모델 38개에 관한 연비 데이터 `mpg` 이용

- 연도별(`year`) 7개 차종(`class`)의 빈도는 어떻게 되는가?

```{r}
mpg |> 
    select(manufacturer:model, year, class) |> 
    str()
```


---

## .brand-blue[흔히들 하는 시각화]

.pull-left[


```{r mpg1, warning = FALSE, fig.height = 6.5, fig.retina = 2, fig.cap = "그림 1.1", echo = FALSE}
mpg %>% 
  ggplot(aes(x = factor(year), fill = class)) +
  geom_bar() +
  labs(x = "year")
```

]

--

.pull-right[

```{r, eval = FALSE}
mpg %>% 
  ggplot(aes(x = factor(year), 
             fill = class)) +
  geom_bar() +
  labs(x = "year")
```

]

---

## .brand-blue[음영 활용하기 I - Preprocessing]

.left-column[
]
.right-column[

```{r}
# Group classes into three groups (to reduce colors to 3)
dat <- mpg %>% 
  mutate(
    year = factor(year),
    class_group = case_when(
      class %in% c('2seater', 'compact', 'midsize') ~ "grp1",
      class == 'minivan' ~ "grp2",
      T ~ "grp3"
    )
  )
```

]

---

## .brand-blue[음영 활용하기 I - Visualization]

.pull-left[

```{r mpg2, warning = FALSE, fig.height=6.5, fig.retina = 2, fig.cap = "그림 1.2", echo = FALSE}
shades_plt <- dat %>% 
  ggplot(aes(x = year, 
             fill = class_group, 
             alpha = class)) +
  geom_bar() +
  labs(
    x = 'Year',
    y = 'Counts',
    alpha = 'Class',
    title = 'Show shades, not hues'
  )
shades_plt 
```

]

--

.pull-right[

```{r, eval = FALSE}
shades_plt <- dat %>% 
  ggplot(aes(x = year, 
             fill = class_group, #<<
             alpha = class)) + #<<
  geom_bar() +
  labs(
    x = 'Year',
    y = 'Counts',
    alpha = 'Class',
    title = 'Show shades, not hues'
  )
shades_plt 
```

]

---

## .brand-blue[음영 활용하기 II - Adjusting]

.left-column[
]
.right-column[

```{r}
# Color-blind safe colors
colors <-  thematic::okabe_ito(3)
# Possible levels of transparency (one for each class)
alpha_max <- 1
alpha_min <- 0.7
alpha_vals <- c(
  seq(alpha_max, alpha_min, length.out = 4), 
  seq(alpha_min, alpha_max, length.out = 4)[-1]
)
alpha_vals
```

]

---

## .brand-blue[음영 활용하기 II - Visualization]

.pull-left[

```{r mpg3, warning = FALSE, fig.height=6.5, fig.retina = 2, fig.cap = "그림 1.3", echo = FALSE}
# Tweak previous plot
shades_plt <- shades_plt +
  scale_fill_manual(values = colors) +
  scale_alpha_manual(values = alpha_vals)
shades_plt
```

]

--

.pull-right[

```{r, eval = FALSE}
# Tweak previous plot
shades_plt <- shades_plt +
  scale_fill_manual(values = colors) + #<<
  scale_alpha_manual(values = alpha_vals) #<<
shades_plt
```

]

---


## .brand-blue[음영 활용하기 III - Adjusting and visualization]

.pull-left[

```{r mpg4, warning = FALSE, fig.height=6.5, fig.retina = 2, fig.cap = "그림 1.4", echo = FALSE}
# Tweak previous plot
shades_plt <- shades_plt +
  guides(
    fill = guide_none(),
    alpha = guide_legend(
      override.aes = list(fill = colors[c(1, 1, 1, 2, 3, 3, 3)]
      )
    )
  ) 
shades_plt
```

]

--

.pull-right[

- `guides()` 이용

```{r, eval = FALSE}
shades_plt <- shades_plt +
  guides( #<<
    fill = guide_none(), #<<
    alpha = guide_legend( #<<
      override.aes = #<<
          list(fill = colors[c(1, 1, 1, 2, 3, 3, 3)]) #<<
    ) #<<
  ) #<<
shades_plt
```

]

---

class: inverse, middle, center

# 음영만으로는 부족해

---

## .brand-blue[컬러블록에 줄 추가하기]

.pull-left[

```{r mpg5, warning = FALSE, fig.height=6.5, fig.retina = 2, fig.cap = "그림 1.5", echo = FALSE}
dat %>% 
  ggplot(aes(x = year, fill = class_group, alpha = class)) +
  geom_bar(col = 'white') + # Add lines for distinction
  scale_fill_manual(values = colors) +
  scale_alpha_manual(values = alpha_vals) +
  guides(
    fill = guide_none(),
    alpha = guide_legend(override.aes = list(fill = colors[c(1, 1, 1, 2, 3, 3, 3)]))
  ) +
  labs(
    x = 'Year',
    y = 'Counts',
    alpha = 'Class',
    title = 'Group categories together by color, \nbut keep showing them'
  )
```

]

--

.pull-right[

```{r, eval = FALSE}
dat %>% 
  ggplot(aes(x = year, fill = class_group, alpha = class)) +
  geom_bar(col = 'white') + #<<
  scale_fill_manual(values = colors) +
  scale_alpha_manual(values = alpha_vals) +
  guides(
    fill = guide_none(),
    alpha = guide_legend(override.aes = list(fill = colors[c(1, 1, 1, 2, 3, 3, 3)]))
  ) +
  labs(
    x = 'Year', y = 'Counts', alpha = 'Class',
    title = 'Group categories together by color, \nbut keep showing them'
  )
```

]

---

class: inverse, middle, center

# 전달하고 싶은 내용만 강조하자

---

## .brand-blue[스파게티 플롯 예제]

- [Our World in Data](https://ourworldindata.org/time-with-others-lifetime)의 미국인 설문조사 자료 이용

- 우리는 일생동안 누구(`person`)와 시간(`minutes`)을 보내는가?

.pull-left[
```{r,  message = FALSE}
time_data <- read_csv("./time-spent-with-relationships-by-age-us.csv") %>% 
  rename_with(
    ~c('Entitity', 'Code', 'Age', 
       'alone', 'friends', 'children', 
       'parents', 'partner', 'coworkers')
  ) %>% 
  pivot_longer(
    cols = alone:coworkers, names_to = 'person',
    values_to = 'minutes'
  ) %>% 
  janitor::clean_names() %>% 
  filter(age <= 80)
```
]

.pull-right[
```{r}
str(time_data)
```

]

---
## .brand-blue[흔히들 하는 시각화]

.pull-left[

```{r spend1, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 2.1", echo = FALSE}
# Color-blind safe colors
colors <- thematic::okabe_ito(7)[-6]

# Line plot
p <- time_data %>% 
  ggplot(aes(x = age, y = minutes, col = person)) +
  geom_line(size = 1.5) +
  scale_color_manual(values = colors) +
  coord_cartesian(xlim = c(15, 81), expand = F) +
  scale_y_continuous(minor_breaks = NULL) +
  labs(x = 'Age (in years)', y = 'Minutes', col = 'Time spent')
p
```

]

--

.pull-right[

```{r, eval = FALSE}
# Color-blind safe colors
colors <- thematic::okabe_ito(7)[-6]

# Line plot
p <- time_data %>% 
  ggplot(aes(x = age, y = minutes, col = person)) +
  geom_line(size = 1.5) +
  scale_color_manual(values = colors) +
  coord_cartesian(xlim = c(15, 81), expand = F) +
  scale_y_continuous(minor_breaks = NULL) +
  labs(x = 'Age (in years)', y = 'Minutes', col = 'Time spent')
p
```

]
---
## .brand-blue[전달력 높이기 I]


.pull-left[

```{r spend2, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 2.2", echo = FALSE}
library(gghighlight)
alone_plt <- p + 
  gghighlight(person == 'alone', use_direct_label = F) +
  labs(title = 'Emphasize just one or a few categories')
alone_plt
```

]

--

.pull-right[

- `gghighlight()` 이용

```{r, eval = FALSE}
library(gghighlight)
alone_plt <- p + 
  gghighlight(person == 'alone', #<<
              use_direct_label = F) + #<<
  labs(title = 'Emphasize just one or a few categories')
alone_plt
```

]

---

## .brand-blue[전달력 높이기 II]

.pull-left[

```{r spend_add, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 2.3", echo = FALSE}
alone_plt +
  annotate(
    'text',
    x = 15,
    y = 455,
    label = 'We spend a lot of time alone...',
    hjust = 0,
    vjust = 0,
    family = 'firasans',
    size = 7
  )
```

]

--

.pull-right[

```{r, eval = FALSE}
alone_plt +
  annotate(
    'text',
    x = 15,
    y = 455,
    label = 'We spend a lot of time alone...',
    hjust = 0,
    vjust = 0,
    family = 'firasans',
    size = 7
  )
```

]

---

## .brand-blue[전달력 높이기 III]

```{r spend3, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 2.4", echo = FALSE}
age_40_plt <- p + 
  gghighlight(
    person %in% c('alone', 'children'), 
    age >= 38, 
    use_direct_label = F
  ) +
  geom_segment(x = 38, xend = 38, y = -Inf, yend = 300, linetype = 2, col = 'grey20') +
  labs(title = 'Emphasize just one or a few categories') 

age_40_plt +
  annotate(
    'text',
    x = 15,
    y = 403,
    label = 'Around the age of 40, we spend \nless time with children and \nmore time alone.',
    hjust = 0,
    vjust = 0,
    family = 'firasans',
    lineheight = 0.85,
    size = 5.5
  )
```

---
## .brand-blue[전달력 높이기 III]

- `ggphighlight()`를 이용해 다중 하이라이팅

```{r, eval = FALSE}
age_40_plt <- p + 
  gghighlight(person %in% c('alone', 'children'), age >= 38,  use_direct_label = F) + #<<
  geom_segment(x = 38, xend = 38, y = -Inf, yend = 300, linetype = 2, col = 'grey20') +
  labs(title = 'Emphasize just one or a few categories') 

age_40_plt +
  annotate(
    'text', x = 15, y = 403,
    label = 'Around the age of 40, we spend \nless time with children and \nmore time alone.',
    hjust = 0, vjust = 0,
    family = 'firasans',
    lineheight = 0.85,
    size = 5.5
  )
```

---

class: inverse, middle, center

# 라벨링 활용하기

---

## .brand-blue[라벨링 활용하기 I]

.pull-left[

```{r spend4, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 3.1", echo = FALSE}
alone_plt +
  annotate(
    'text', x = 15, y = 455,
    label = 'We spend a lot of time alone...',
    hjust = 0, vjust = 0,
    family = 'firasans',
    size = 7) +
  annotate(
    'text',  x = 70,  y = 420, 
    label = 'alone',
    hjust = 0, vjust = 0, size = 7,
    family = 'firasans', color = colors[1]) +
  labs(title = 'Label directly') +
  theme(legend.position = 'none')
```

]

--

.pull-right[

```{r, eval = FALSE}
alone_plt +
  annotate(
    'text', x = 15, y = 455,
    label = 'We spend a lot of time alone...',
    hjust = 0, vjust = 0,
    family = 'firasans',
    size = 7) +
  annotate( #<<
    'text',  x = 70,  y = 420, #<<
    label = 'alone', #<<
    hjust = 0, vjust = 0, size = 7, #<<
    family = 'firasans', color = colors[1]) + #<<
  labs(title = 'Label directly') +
  theme(legend.position = 'none') #<<
```

]

---

## .brand-blue[라벨링 활용하기 II - Adjusting]

- `ggtext` 패키지를 통해 richtext로 변환 및 HTML 문법 활용

```{r}
library(ggtext)
color_alone <- glue::glue(
  "We spend a lot of time <span style = 'color:{colors[1]};'>alone</span>..." #<<
)
color_alone
```

---

## .brand-blue[라벨링 활용하기 II - Visualization]

.pull-left[

```{r spend5, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 3.1", echo = FALSE}
alone_plt +
  labs(title = 'Label directly') +
  annotate(
    'richtext',
    x = 15,
    y = 400,
    label = color_alone,
    hjust = 0, vjust = 0,
    family = 'firasans',
    size = 6,
    label.color = NA
  ) +
  theme(legend.position = 'none')
```   

]

--

.pull-right[

```{r, eval = FALSE}
alone_plt +
  labs(title = 'Label directly') +
  annotate(
    'richtext', #<<
    x = 15,
    y = 400,
    label = color_alone, #<<
    hjust = 0, vjust = 0,
    family = 'firasans',
    size = 6,
    label.color = NA #<<
  ) +
  theme(legend.position = 'none') #<<
```

]

---

## .brand-blue[라벨링 활용하기 II - Adjusting]

```{r}
age_40_text <- glue::glue(
  "Around the age of 40, we spent <br> less time 
  with <span style = 'color:{colors[2]};'>children</span> 
  and <br> more time <span style = 'color:{colors[1]};'>alone</span>."
)
age_40_text
```

---

## .brand-blue[라벨링 활용하기 II - Visualization]

.pull-left[

```{r spend6, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 3.2", echo = FALSE}
age_40_plt +
  labs(title = 'Label directly') +
  annotate(
    'richtext',
    x = 15, y = 350, 
    label = age_40_text,
    hjust = 0, vjust = 0,
    family = 'firasans',
    lineheight = 1.25,
    size = 5,
    label.color = NA
  ) +
  theme(legend.position = 'none')
``` 

]

--

.pull-right[
  
```{r, eval = FALSE}
age_40_plt +
  labs(title = 'Label directly') +
  annotate(
    'richtext', #<<
    x = 15, y = 350, 
    label = age_40_text, #<<
    hjust = 0, vjust = 0,
    family = 'firasans',
    lineheight = 1.25,
    size = 5,
    label.color = NA #<<
  ) +
  theme(legend.position = 'none') #<<
```
  
]

---

## .brand-blue[라벨링 활용하기 III - Preprocessing]

- 빈도 계산

.pull-left[
```{r, eval = FALSE}
manual_counts <- mpg %>% 
  count(year, class) %>% 
  mutate(
    year = factor(year),
    class_group = case_when(
      class %in% c('2seater', 'compact', 'midsize') ~ "grp1",
      class == 'minivan' ~ "grp2",
      T ~ "grp3"
    )
  ) 
head(manual_counts)
```
]

.pull-right[
```{r, echo = FALSE}
manual_counts <- mpg %>% 
  count(year, class) %>% 
  mutate(
    year = factor(year),
    class_group = case_when(
      class %in% c('2seater', 'compact', 'midsize') ~ "grp1",
      class == 'minivan' ~ "grp2",
      T ~ "grp3"
    )
  ) 
head(manual_counts)
```  
]

---


## .brand-blue[라벨링 활용하기 III - Preprocessing]

- 레이블 높이 계산

.pull-left[
```{r, eval = FALSE}
labels <- manual_counts %>% 
  mutate(class = factor(class)) %>%  
  group_by(year) %>% 
  arrange(year, desc(class)) %>% 
  mutate(
    csum = cumsum(n),  #<<
    n = (lag(csum, default = 0) + csum) / 2 #<<
  )
head(labels)
```
]

.pull-right[
```{r, echo = FALSE}
labels <- manual_counts %>% 
  mutate(class = factor(class)) %>%  
  group_by(year) %>% 
  arrange(year, desc(class)) %>% 
  mutate(
    csum = cumsum(n), 
    n = (lag(csum, default = 0) + csum) / 2
  )
head(labels)
```  
]

---

## .brand-blue[라벨링 활용하기 III - Visualization]

.pull-left[
```{r mpg-add, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 3.3", echo = FALSE}
shades_plt2 <- manual_counts %>%
  ggplot(aes(x = year, y = n, fill = class_group)) +
  geom_col(aes(alpha = class), col = 'white') + #<<
  scale_fill_manual(values = colors) +
  scale_alpha_manual(values = alpha_vals) +
  labs(
    x = 'Year', y = 'Counts',
    alpha = 'Class', title = 'Label directly'
  )
shades_plt2
```
]

.pull-right[
```{r, eval = FALSE}
shades_plt2 <- manual_counts %>% #<<
  ggplot(aes(x = year, y = n, fill = class_group)) +
  geom_col(aes(alpha = class), col = 'white') + #<<
  scale_fill_manual(values = colors) +
  scale_alpha_manual(values = alpha_vals) +
  labs(
    x = 'Year', y = 'Counts',
    alpha = 'Class', title = 'Label directly'
  )
shades_plt2
```
]

---

## .brand-blue[라벨링 활용하기 III - Visualization]

```{r mpg6, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 3.4", echo = FALSE}
shades_plt2 +
  # Add all but one label
  geom_text(
    data = labels %>% filter(class != '2seater'), aes(label = class), #<<
    col = 'white', family = 'firasans', size = 5, fontface = 'bold'
  ) +
  # Add 2seater label
  geom_text(
    data = labels %>% filter(class == '2seater'), aes(y = n + 3, label = class), #<<
    col = 'black', family = 'firasans', size = 5, fontface = 'bold'
  ) +
  theme(legend.position = 'none')
```

---

## .brand-blue[라벨링 활용하기 III - Visualization]


```{r, warning = FALSE, fig.height = 6, fig.retina = 2, fig.cap = "그림 3.2", eval = FALSE}
shades_plt2 +
  # Add all but one label
  geom_text(
    data = labels %>% filter(class != '2seater'), aes(label = class), #<<
    col = 'white', family = 'firasans', size = 5, fontface = 'bold'
  ) +
  # Add 2seater label
  geom_text(
    data = labels %>% filter(class == '2seater'), aes(y = n + 3, label = class), #<<
    col = 'black', family = 'firasans', size = 5, fontface = 'bold'
  ) +
  theme(legend.position = 'none')
```

---

## .brand-blue[Before & After - Bar chart]

.pull-left[
```{r before-1, warning = FALSE, fig.height=6.5, fig.retina = 2, echo = FALSE}
mpg %>% 
  ggplot(aes(x = factor(year), 
             fill = class)) +
  geom_bar() +
  labs(x = "year")
```
]

--

.pull-right[
```{r after-1, warning = FALSE, fig.height = 6.5, fig.retina = 2, echo = FALSE}
shades_plt2 +
  # Add all but one label
  geom_text(
    data = labels %>% filter(class != '2seater'), aes(label = class), #<<
    col = 'white', family = 'firasans', size = 5, fontface = 'bold'
  ) +
  # Add 2seater label
  geom_text(
    data = labels %>% filter(class == '2seater'), aes(y = n + 3, label = class), #<<
    col = 'black', family = 'firasans', size = 5, fontface = 'bold'
  ) +
  theme(legend.position = 'none') +
  labs(title = "")
``` 
]

---

## .brand-blue[Before & After - Spaghetti plot]

.pull-left[
```{r before-2, warning = FALSE, fig.height = 6, fig.retina = 2, echo = FALSE}
p
```
]

--

.pull-right[
```{r after-2, warning = FALSE, fig.height = 6, fig.retina = 2, echo = FALSE}
alone_plt +
  annotate(
    'richtext', #<<
    x = 15,
    y = 400,
    label = color_alone, #<<
    hjust = 0, vjust = 0,
    family = 'firasans',
    size = 6,
    label.color = NA #<<
  ) +
  theme(legend.position = 'none') + #<<
  labs(title = "")
``` 
]


---

## .brand-blue[Before & After - Spaghetti plot]

.pull-left[
```{r before-3, warning = FALSE, fig.height = 6, fig.retina = 2, echo = FALSE}
p
```
]

--

.pull-right[
```{r after-3, warning = FALSE, fig.height = 6, fig.retina = 2, echo = FALSE}
age_40_plt +
  annotate(
    'richtext', #<<
    x = 15, y = 350, 
    label = age_40_text, #<<
    hjust = 0, vjust = 0,
    family = 'firasans',
    lineheight = 1.25,
    size = 5,
    label.color = NA #<<
  ) +
  theme(legend.position = 'none') +
  labs(title = "")
``` 
]

---
class: inverse, middle, center

# References

---

## .brand-blue[References]

[1] Rapp, Albert. “4 Ways to Use Colors in Ggplot More Efficiently.” Blog. Albert Rapp, February 19, 2022. https://albert-rapp.de/posts/ggplot2-tips/07_four_ways_colors_more_efficiently/07_four_ways_colors_more_efficiently.html.

### .brand-blue[Theme and CSS in xaringan slides]

- https://github.com/statisticsplaybook/statisticsplaybook

- https://github.com/garthtarr/sydney_xaringan

---

class: inverse

# Thanks!

.pull-right[.pull-down[

<a href="mailto:favorite@kakao.com">
.white[`r icons::fontawesome("paper-plane")` favorite@kakao.com]
</a>

<a href="https://github.com/be-favorite">
.white[`r icons::fontawesome("github")` @be-favorite]
</a>

<a href="https://twitter.com/TaemoBang">
.white[`r icons::fontawesome("twitter")` @TaemoBang]
</a>

<a href="https://www.taemobang.com">
.white[`r icons::fontawesome("link")` Blog]
</a>

<br><br><br>

]]